Seismic parameter estimation and the Canadian crust

### 2013-06-14 paper/draftBP.tex
draftBP

  Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 km^2. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

@@

%% -----------------------------------------------%%
\section{Data and methods}
  Analysis of bulk Canadian continental crust requires estimates of crustal properties such as seismic-wave velocity and crustal thickness. This study draws on three sources for these crustal properties. The primary source of estimates are data computed by first processing raw seismograms into receiver functions and then inverting the RF's for parameter estimates. These primary data are supplemented and qualified by two existing datasets, the statistically compiled Crust 2.0 dataset and a compilation of pre-processed active source data.

@@

  The primary data utilized in this study are computed from teleseismic P-wave seismograms representing discrete seismic events. These events are comprised by the seismic records of more than 700 earthquake sources occuring between the years 2000 and 2012 at 343 broadband seismic stations across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Events are filtered by the epicentral distance from source to receiver with only the events within a 30 to 100 degree window being included. Seismic events are further filtered by hand picking those with reasonable signal to noise ratio and of sufficient impulsiveness that arrival times can be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

Seismograms are selected between the years 2000 and 2012 which employ more 1000 earthquake sources. All teleseismic events outside of the 30 to 100 degrees of epicentral distance are removed before processing. Data is selected for reasonable signal to noise ratio and sufficient impulsiveness that arrival times can be calculated. All data is analyzed using receiver function methods, an approach which has been widely employed over many years to investigate the earths structure.

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{thompsonComparisonR}
  \caption{Comparison of $\frac{V_P}{V_R}$ data from this study with data from Thomspson et. al. (2010). Data shows a Pearson correlation of 0.5}
  \label{fig:thompsonCompR}
\end{figure}

>>>>

  Seismic studies of the Canadian continental crust predominately analyze specific geological features or regions rather than taking a comprehensive and comparative inter-regional analysis. The primary reason for this is the poor resolution afforded by the seismic networks currently and historically deployed across Canada. The Canadian continental landmass is composed of at least fifteen large geological provinces as recognized by the Geological Survey of Canada (GSC). Each of these regions is itself complex and heterogeneous and often larger than most European nations. On top of this there is poor seismic coverage, roughly one seismic station per 25,000 $km^2$. Many of these stations are clustered near areas of geologic interest such as the Cascadia Subduction zone or population and research centres such as Southern Ontario or areas of significant resource interest like the diamondiferous Great Slave Lake area. This leaves vast areas of the Canadian landmass completely unsampled. Despite these limitations, a low resolution comprehensive and comparative study is still feasible for some of the geological regions comprising the Canadian continental crust. Direct comparison between the bulk average seismic properties of geological regions and between aggregated regions and global averages provide some new insight into the crustal composition of the Canadian landmass. The accumulated dataset also affords investigations into variations of bulk crustal parameters such as Poisson's Ratio and crustal thickness with age and tectonic environment as well as some basic statistical data on interesting crustal features.

@@

%% -----------------------------------------------%%
\section{Data and methods}
  Analysis of bulk Canadian continental crust requires estimates of crustal properties such as seismic-wave velocity and crustal thickness. This study draws on three sources for these crustal properties. The first and primary source for these estimates are data computed by processing raw seismograms into receiver functions and then inverting these RF's for parameter estimates. The other two datasets, the statistically compiled Crust 2.0 dataset and a compilation of pre-processed active source data, are used to qualify and compare with the primary dataset.

@@

  The primary data utilized in this study are computed from teleseismic P-wave seismograms representing discrete seismic events. These events are comprised of seismic records representing more than 700 earthquake sources occuring between the years 2000 and 2012 at 343 broadband seismic stations across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Events are filtered by the epicentral distance from source to receiver with only the events within a 30 to 100 degree window being included. Seismic events are further filtered by hand picking those with reasonable signal to noise ratio and of sufficient impulsiveness that arrival times can be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

  The first stage of processing requires the transformation of teleseismic data into receiver functions. In its generic form this transformation involves deconvolving an approximation of the source signal from channels representing ground motion. The resulting waveform contains discrete pulses which corrispond to arrivals of S-wave energy scattered from subsurface discontinuities. The resolution of these pulses has first-order effectsTo gain better resolution of these peaks we follow a wave field decomposition approach to To better focus this energy the North and East channels are rotated into radial and transverse components. Then the radial and vertical components are then decomposed into P, SV and SH components with a wave field decomposition transfer matrix [Bostock, 1998].

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{thompsonComparisonR}
  \caption{Comparison of $V_P / V_R$ data from this study with data from Thomspson et. al. (2010). Data shows a Pearson correlation of 0.5}
  \label{fig:thompsonCompR}
\end{figure}
